% Appendix A. Reproducibility checklist. To be \input{} into the paper or compiled separately as supplementary.

\appendix
\section{Reproducibility Appendix}
\label{app:repro}

\subsection{Pinned artifacts}

All numeric claims in this paper are computed from the following pinned
data and code:

\begin{itemize}
\item \textbf{HGNC dictionary.} 3-column TSV (Approved symbol, Previous
symbols, Alias symbols), 50{,}094 rows, mtime 2026-01-11 02:06:29 UTC,
SHA-256
\texttt{b9482a5247a4162017bd9731ac6300757b9b7511ec9a02f99bafd6b4a43a89fc}.
\item \textbf{BEELINE reference networks.} Archive
\texttt{BEELINE-Networks.zip}, 16 MB, SHA-256
\texttt{d54fc6ed6529141e85f94803d2c730a8173a3e867026227528ff1ae832765607},
from Zenodo record 3701939. Contains
\texttt{human-tfs.csv} (1{,}563 entries),
\texttt{Networks/human/hESC-ChIP-seq-network.csv} (436{,}563 edges),
\texttt{Networks/human/HepG2-ChIP-seq-network.csv} (342{,}862 edges),
\texttt{Networks/human/Non-specific-ChIP-seq-network.csv}, and
\texttt{Networks/human/STRING-network.csv}.
\item \textbf{BEELINE scRNA-seq inputs.} Archive
\texttt{BEELINE-data.zip}, 250 MB, SHA-256
\texttt{1fc849499492c25ac72514e79a339c39aa057e58cd272097450875cecb380afa},
from the same Zenodo record. We use
\texttt{inputs/scRNA-Seq/hESC/ExpressionData.csv} (17{,}735 genes
$\times$ 758 cells, log-normalized) and
\texttt{inputs/scRNA-Seq/hHep/ExpressionData.csv} (11{,}515 genes
$\times$ 425 cells, log-normalized).
\item \textbf{scGPT checkpoint.} Whole-human checkpoint from
\cite{cui2024scgpt}, 196 MB,
\texttt{external/scGPT\_checkpoints/whole-human/best\_model.pt}.
Vocabulary size 60{,}697; max\_seq\_len 1\,200; 12 layers; 8 heads.
\item \textbf{Geneformer checkpoint.} Geneformer-V1-10M from
\cite{theodoris2023geneformer}, HuggingFace
\texttt{ctheodoris/Geneformer} commit
\texttt{fcd26c45fc30fba1989e586bdc46bc366dda8655}, 39 MB. Vocabulary
size 25{,}426; max\_position\_embeddings 2\,048; 6 layers; 4 heads.
\end{itemize}

\subsection{Procedure}

\begin{enumerate}
\item Run \texttt{scripts/prep\_beeline.py --tag hESC}. Inverts BEELINE
log-normalization to integer counts, applies the pinned HGNC mapping,
emits \texttt{outputs/prep/hESC/hESC\_raw.h5ad} and
\texttt{hESC\_universe.json}. Repeat for hHep.
\item Run \texttt{scripts/run\_baselines.py --tag hESC --method
\{pearson,spearman,random,genie3,grnboost2\}}. Each emits a Parquet
\texttt{outputs/baselines/hESC/<method>.parquet} with columns
\texttt{(source, target, score)} covering exactly $U(\text{hESC})$.
\item Run \texttt{scripts/run\_scgpt\_attention.py --tag hESC
--max-genes 1000 --device cpu --batch-size 2}. Emits six aggregation
variants into \texttt{outputs/scgpt/hESC/}.
\item Run \texttt{scripts/run\_geneformer\_attention.py --tag hESC}.
Emits six aggregation variants into \texttt{outputs/geneformer/hESC/}.
\item Run \texttt{scripts/eval\_universe\_aware.py --tag hESC} and
\texttt{--tag hHep}. Emits
\texttt{outputs/eval/<tag>\_metrics.parquet}.
\item Run \texttt{scripts/permutation\_null.py --tag <tag>
--n-perm 1000 --n-boot 2000}. Emits
\texttt{outputs/eval/<tag>\_permutation.parquet} and
\texttt{<tag>\_paired\_bootstrap.parquet}.
\item Run \texttt{scripts/variance\_decomp.py --y log\_aupr}. Emits
\texttt{outputs/eval/variance\_decomp\_log\_aupr.csv}.
\item Run \texttt{scripts/format\_results.py}. Emits
\texttt{outputs/eval/paper\_snippets.md} which contains every numeric
value referenced in this paper.
\end{enumerate}

\subsection{Seeds and randomness}

All randomized procedures use the seeded interfaces in
\texttt{numpy.random.default\_rng(seed)} and
\texttt{sklearn.ensemble.RandomForestRegressor(random\_state=seed)} /
\texttt{GradientBoostingRegressor(random\_state=seed)}. Per-target
regressor seeds are derived as \texttt{base\_seed + target\_index}.
Permutation seeds use a single \texttt{rng = np.random.default\_rng(42)}.
Paired-bootstrap seeds use the same \texttt{rng}.

\subsection{Environment}

\begin{itemize}
\item Python 3.11.9 (Anaconda)
\item PyTorch 2.11 (MPS backend)
\item Transformers (current pinned version in lockfile)
\item scanpy 1.10.2, anndata 0.12.7
\item arboreto 0.1.6 (with sklearn-only replacement for the
GENIE3 / GRNBoost2 fits; see \texttt{scripts/run\_baselines.py}
\texttt{run\_arboreto()} for the equivalent fits via
\texttt{sklearn.ensemble} + \texttt{joblib})
\item statsmodels for the ANOVA decomposition
\end{itemize}

\subsection{Hardware used in our runs}

Apple M-series with 10 CPU cores, MPS GPU, 32 GB RAM. scGPT extraction
runs on CPU with \texttt{OMP\_NUM\_THREADS=2} and
\texttt{MKL\_NUM\_THREADS=2} (flash attention is not available on this
platform). Geneformer-V1-10M runs on MPS. Total wall time for the
hESC+hHep pipeline including 1\,000-permutation null and
2\,000-paired-bootstrap is approximately \texttt{TBA hours}.
